---
title: "Diagnostics Metrics"
author: "Anna Lizarov"
date: "November 29, 2018"
output: html_document
---

In the following project, the data from an one level of an online geography tutoring system used by 5th grade students will be inspected. The game involves a pre-test of geography knowledge (pre.test), a series of assignments for which you have the average score (av.assignment.score),  the number of messages sent by each student to other students about the assignments (messages), the number of forum posts students posted asking questions about the assignment (forum.posts), a post test at the end of the level (post.test) and whether or not the system allowed the students to go on to the next level (level.up).  

## Part I

#Upload data
```{r}
D1 <- read.csv("online.data.csv", header=TRUE)
```

```{r}
library(dplyr)
library(tidyr)
D2 <- D1[,-1]
D3 <- D2 %>% mutate(level.up=ifelse(level.up=="yes",1,0)) 
D4 <- D3 %>% gather("Variables","Value", 1:6)
```

#Visualization 
```{r}
#Start by creating histograms of the distributions for all variables (#HINT: look up "facet" in the ggplot documentation)

library(ggplot2)
ggplot(D4, aes(Value))+
  geom_histogram()+
  facet_wrap(~Variables, scale="free")
```

```{r}
#Then visualize the relationships between variables
library(corrplot)
D5 <- D1[,2:6]
COR <- cor(D5)

corrplot(COR, order="AOE", method="circle", tl.pos="lt", type="upper",        
tl.col="black", tl.cex=0.6, tl.srt=45, 
        addCoef.col="black", addCoefasPercent = TRUE,
        sig.level=0.50, insig = "blank")

#Try to capture an intution about the data and the relationships
```

```{r}
#Scatterplot 
pairs(D5)
```


```{r}
#Answer: There is strong positive relationship between post test score and the number of messages sent by each student to other students about the assignments. However, there is a weak positive relationship between post test score and the number of forum posts students posted asking questions about the assignment. This indicates that messaging is more effective in improving grades for geography than forum posts. It is important to note that, based on the histogram, students are more actively engaged in messaging than posting questions on a forum regarding the assignment. Nonetheless, the average assignment score is moderately to highly related to post test score and messages. 
```

#Classification tree
```{r}
#Create a classification tree that predicts whether a student "levels up" in the online course using three variables of your choice (As we did last time, set all controls to their minimums)
library(rpart)
rp<- rpart(level.up~forum.posts+post.test.score+messages, method="class", data=D3)
printcp(rp)
#Plot and generate a CP table for your tree 
post(rp, file = "tree1.ps", title = "Online Geography Tutoring System") #This creates a pdf image of the tree

```

```{r}
#Generate a probability value that represents the probability that a student levels up based your classification tree 

D3$pred <- predict(rp, type = "prob")[,2]#Last class we used type = "class" which predicted the classification for us, this time we are using type = "prob" to see the probability that our classififcation is based on.
```

## Part II
#Now you can generate the ROC curve for your model. You will need to install the package ROCR to do this.
```{r}
library(ROCR)

#Plot the curve
pred.detail <- prediction(D3$pred, D3$level.up) 
plot(performance(pred.detail, "tpr", "fpr"))
abline(0, 1, lty = 2)
```

```{r}
#Calculate the Area Under the Curve
unlist(slot(performance(pred.detail,"auc"), "y.values"))#Unlist liberates the AUC value from the "performance" object created by ROCR     # AUC  = 0.855
```

```{r}
#Now repeat this process, but using the variables you did not use for the previous model and compare the plots & results of your two models. Which one do you think was the better model? Why?
rp2<- rpart(level.up~av.assignment.score+pre.test.score, method="class", data=D3)
printcp(rp2)
#Plot and generate a CP table for your tree 
post(rp2, file = "tree2.ps", title = "Online Geography Tutoring System 2") #This creates a pdf image of the tree
```

```{r}
D3$pred2 <- predict(rp2, type = "prob")[,2]

#Plot the curve
pred2.detail <- prediction(D3$pred2, D3$level.up) 
plot(performance(pred2.detail, "tpr", "fpr"))
abline(0, 1, lty = 2)
```

```{r}
#Calculate the Area Under the Curve
unlist(slot(performance(pred2.detail,"auc"),"y.values"))    #AUC = 0.977
```


```{r}
#Answer: Model 2 is better since its area under the ROC (AUC) is greater than that of model 1. In particular, AUC for model 2 is 0.977 whereas the AUC for model 1 is 0.855. This suggests that model 2 is better at predicting that the system will allow the student to go to the next level since it has a higher true positive rate (sensitivity). 
```

## Part III
#Thresholds
```{r}
#Look at the ROC plot for your first model. Based on this plot choose a probability threshold that balances capturing the most correct predictions against false positives. Then generate a new variable in your data set that classifies each student according to your chosen threshold.

#Model 1:

threshold.pred1 <- 0.70
D3$threshold.pred1 <- ifelse(D3$pred <= threshold.pred1, 0,1) 

D3.a <- count(D3, level.up, threshold.pred1)
D3.a$outcome <- ifelse(D3.a$level.up== 1 & D3.a$threshold.pred1== 1, "TP", ifelse(D3.a$level.up== 1 & D3.a$threshold.pred1== 0, "FN", ifelse(D3.a$level.up== 0 & D3.a$threshold.pred1== 1, "FP", "TN")))
D3.a
```

```{r}
#Now generate three diagnostics:

D3$accuracy.model1 <- ((D3.a[D3.a$outcome=="TP",]$n + D3.a[D3.a$outcome=="TN",]$n) / sum(D3.a$n))  
 # = 0.841

D3$precision.model1 <- D3.a[D3.a$outcome=="TP",]$n/(D3.a[D3.a$outcome=="TP",]$n+D3.a[D3.a$outcome=="FP",]$n)     # = 0.742

D3$recall.model1 <- D3.a[D3.a$outcome=="TP",]$n/(D3.a[D3.a$outcome=="TP",]$n+D3.a[D3.a$outcome=="FN",]$n)        # = 0.923

```


```{r}
#First generate the table of comparisons
table1 <- table(D1$level.up, D3$threshold.pred1)
colnames(table1) <- c("no", "yes")
table1
```
```{r}
#Convert to matrix
matrix1 <- as.matrix(table1)

#Calculate kappa
po <-  ((D3.a[D3.a$outcome=="TP",]$n + D3.a[D3.a$outcome=="TN",]$n) / sum(D3.a$n))
pe <- (((D3.a[D3.a$outcome=="TN",]$n+D3.a[D3.a$outcome=="FN",]$n)/sum(D3.a$n))*((D3.a[D3.a$outcome=="TN",]$n+D3.a[D3.a$outcome=="FP",]$n)/sum(D3.a$n))) + (((D3.a[D3.a$outcome=="TP",]$n+D3.a[D3.a$outcome=="FP",]$n)/sum(D3.a$n))* ((D3.a[D3.a$outcome=="TP",]$n+D3.a[D3.a$outcome=="FN",]$n)/sum(D3.a$n)))
(kappa <- (po-pe)/(1-pe))  # = 0.6816
```
```{r}
#Now choose a different threshold value and repeat these diagnostics. What conclusions can you draw about your two thresholds?
threshold.pred2 <- 0.10
D3$threshold.pred2 <- ifelse(D3$pred <= threshold.pred2, 0,1) 

D3.b <- count(D3, level.up, threshold.pred2)
D3.b$outcome <- ifelse(D3.b$level.up== 1 & D3.b$threshold.pred2== 1, "TP", ifelse(D3.b$level.up== 1 & D3.b$threshold.pred2== 0, "FN", ifelse(D3.b$level.up== 0 & D3.b$threshold.pred2== 1, "FP", "TN")))
D3.b
```
```{r}
#Now generate three diagnostics:

D3$accuracy.model1 <- ((D3.b[D3.b$outcome=="TP",]$n + D3.b[D3.b$outcome=="TN",]$n) / sum(D3.b$n))  
 # = 0.841

D3$precision.model1 <- D3.b[D3.b$outcome=="TP",]$n/(D3.b[D3.b$outcome=="TP",]$n+D3.b[D3.b$outcome=="FP",]$n)     # = 0.742

D3$recall.model1 <- D3.b[D3.b$outcome=="TP",]$n/(D3.b[D3.b$outcome=="TP",]$n+D3.b[D3.a$outcome=="FN",]$n)        # = 0.923
```

```{r}
table2 <- table(D1$level.up, D3$threshold.pred2)
colnames(table2) <- c("no", "yes")
table2
```

```{r}
#Convert to matrix
matrix2 <- as.matrix(table2)

#Calculate kappa
po <-  ((D3.b[D3.b$outcome=="TP",]$n + D3.b[D3.b$outcome=="TN",]$n) / sum(D3.b$n))
pe <- (((D3.a[D3.b$outcome=="TN",]$n+D3.b[D3.b$outcome=="FN",]$n)/sum(D3.b$n))*((D3.b[D3.b$outcome=="TN",]$n+D3.b[D3.b$outcome=="FP",]$n)/sum(D3.b$n))) + (((D3.b[D3.b$outcome=="TP",]$n+D3.b[D3.b$outcome=="FP",]$n)/sum(D3.b$n))* ((D3.b[D3.b$outcome=="TP",]$n+D3.b[D3.b$outcome=="FN",]$n)/sum(D3.b$n)))
(kappa <- (po-pe)/(1-pe))  # = 0.6816
```

```{r}
#Model 2:

threshold.pred3 <- 0.15
D3$threshold.pred3 <- ifelse(D3$pred2 <= threshold.pred3, 0, 1) 

D3.c <- count(D3, level.up, threshold.pred3)
D3.c$outcome <- ifelse(D3.c$level.up== 1 & D3.c$threshold.pred3== 1, "TP", ifelse(D3.c$level.up== 1 & D3.c$threshold.pred3== 0, "FN", ifelse(D3.c$level.up== 0 & D3.c$threshold.pred3== 1, "FP", "TN")))
D3.c
```

```{r}
#Now generate three diagnostics:

D3$accuracy.model2 <- ((D3.c[D3.c$outcome=="TP",]$n + D3.c[D3.c$outcome=="TN",]$n) / sum(D3.c$n))  
 # = 0.972

D3$precision.model2 <- D3.c[D3.c$outcome=="TP",]$n/(D3.c[D3.c$outcome=="TP",]$n+D3.c[D3.c$outcome=="FP",]$n)      # = 0.935

D3$recall.model2 <- D3.c[D3.c$outcome=="TP",]$n/(D3.c[D3.c$outcome=="TP",]$n +0)        
 # = 1
```

```{r}
table3 <- table(D1$level.up, D3$threshold.pred3)
colnames(table3) <- c("no", "yes")
table3
```

```{r}
#Convert to matrix
matrix3 <- as.matrix(table3)

#Calculate kappa
po <-  ((D3.c[D3.c$outcome=="TP",]$n + D3.c[D3.c$outcome=="TN",]$n) / sum(D3.c$n))
pe <- (((D3.c[D3.c$outcome=="TN",]$n+0)/sum(D3.c$n))*((D3.c[D3.c$outcome=="TN",]$n+D3.c[D3.c$outcome=="FP",]$n)/sum(D3.c$n))) + (((D3.c[D3.c$outcome=="TP",]$n+D3.c[D3.c$outcome=="FP",]$n)/sum(D3.c$n))* ((D3.c[D3.c$outcome=="TP",]$n+0)/sum(D3.c$n)))
(kappa <- (po-pe)/(1-pe)) # = 0.942
```

```{r}
# Another Threshold:

threshold.pred3 <- 0.92
D3$threshold.pred3 <- ifelse(D3$pred2 <= threshold.pred3, 0, 1) 

D3.d <- count(D3, level.up, threshold.pred3)
D3.d$outcome <- ifelse(D3.d$level.up== 1 & D3.d$threshold.pred3== 1, "TP", ifelse(D3.d$level.up== 1 & D3.d$threshold.pred3== 0, "FN", ifelse(D3.d$level.up== 0 & D3.d$threshold.pred3== 1, "FP", "TN")))
D3.d
```

```{r}
#Now generate three diagnostics:

D3$accuracy.model2 <- ((D3.d[D3.d$outcome=="TP",]$n + D3.d[D3.d$outcome=="TN",]$n) / sum(D3.d$n))  
 # = 0.972

D3$precision.model2 <- D3.d[D3.d$outcome=="TP",]$n/(D3.d[D3.d$outcome=="TP",]$n+D3.d[D3.d$outcome=="FP",]$n)      # = 0.935

D3$recall.model2 <- D3.d[D3.d$outcome=="TP",]$n/(D3.d[D3.d$outcome=="TP",]$n +0)        
 # = 1
```

```{r}
table4 <- table(D1$level.up, D3$threshold.pred3)
colnames(table4) <- c("no", "yes")
table4
```

```{r}
#Convert to matrix
matrix4 <- as.matrix(table4)

#Calculate kappa
po <-  ((D3.d[D3.d$outcome=="TP",]$n + D3.d[D3.d$outcome=="TN",]$n) / sum(D3.d$n))
pe <- (((D3.d[D3.d$outcome=="TN",]$n+0)/sum(D3.d$n))*((D3.d[D3.d$outcome=="TN",]$n+D3.d[D3.d$outcome=="FP",]$n)/sum(D3.d$n))) + (((D3.d[D3.d$outcome=="TP",]$n+D3.d[D3.d$outcome=="FP",]$n)/sum(D3.d$n))* ((D3.d[D3.d$outcome=="TP",]$n+0)/sum(D3.d$n)))
(kappa <- (po-pe)/(1-pe)) # = 0.942
```

```{r}
# Answer: For both models, different threshold values produced the same results. However, Model 2 is better than Model 1 when it comes to the diagnostic metrics. This indicates that not only model 2 is better at predicting that the system will allow the student to go to the next level, it produces more accurate predictionsand relevant predictions. Likewise, there is a higher reliability betweeen the data and the model.This is indicated by higher accuracy, precision, recall, and kappa values. 

